INLA replication

### 2022-10-18 src/example_inla-replication/inla-replication.Rmd
Clean this up a bit

Consider a three-stage model
\begin{alignat}{2}
  &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \,\mathbf{0},  \mathbf{Q}(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
\end{alignat}
with posterior
\begin{equation}
  p(x, \theta \, | \, y) \propto p(y, x, \theta) = p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta).
\end{equation}

@@

Approximate the posterior marginal $p(\theta \, | \, y) \propto L(\theta)$ using the Laplace approximation $L(\theta) \approx L^\star(\theta)$:
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta)}{\tilde p_G(x \, | \, \theta, y)} \Big\rvert_{x = \mu^\star(\theta)} \equiv L^\star(\theta)
\end{equation}

@@

<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\theta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\theta)) + f(\mu^\star(\theta), \theta)$?* -->

@@

We could also use an adaptive Gauss Hermite quadrature rule.
See the [Implementing Approximate Bayesian Inference using
Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.
## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$

@@

@@

The distribution $p(x \, | \, \theta, y)$ is exact:

@@

Note that the code above in the section marked `(^)` is very similar to that in `blangiardo.cpp`.
Recall that the Laplace approximation $\tilde p(\theta \, | \, y)$, which in this instance coincides with the exact posterior, is given by
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(y, \mu^\star(\theta), \theta)}{\det(Q(\theta))^{1/2}}.
\end{equation}
This form can be seen in `(^)`:

* $p(y, \mu^\star(\theta), \theta)$ corresponds to the calls to `dgamma`, `dnorm` and sum of `dnorm` evaluated at $\mu^\star(\theta)$ which here is simply `par$x_n` (for the Normal distribution the mean of the distribution is the same as the mode).
* The logarithm of $1 / \det(Q(\theta))^{1/2}$ when $\theta$ is a scalar is given by `-0.5 * log(par$tau_n)`.

@@

```{r}
eval_grid <- function(grid = NULL, uniform = FALSE, K = NULL, min = NULL, max = NULL, f) {
  if(uniform) {
    grid <- seq(min, max, length.out = K)
  }
  df <- data.frame(input = grid, output = sapply(grid, f))
  df <- mutate(df, norm_output = output / sum(output))
  return(df)
}

blangiardo_theta <- eval_grid(uniform = TRUE,
                              K = 25, min = 0.001, max = 0.3,
                              f = post_marginal_theta)

dense_theta <- eval_grid(uniform = TRUE,
                         K = 500, min = 0.001, max = 0.3,
                         f = post_marginal_theta)
```

@@

inla_theta <- eval_grid(exp(points), f = post_marginal_theta)

@@

# Different normalising constant?
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y=element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  theme_minimal()
```

@@

```{r}
h <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("x"),
  DLL = "blangiardo",
  hessian = TRUE
)
```

@@

```{r}
f <- MakeADFun(
  data = dat,
  parameters = param,
  DLL = "blangiardo",
  hessian = TRUE
)
```

@@

The mode and comparison (I started by making the mistake of exponentiating `fit$summary.hyperpar$mode` rather than the internal version--on the log scale, as with R and `TMB`--which, because the transformation is non-linear, is not the same).

```{r}
# For some reason the mode in the INLA result below is NA, how to fix this?
kable_data <- data.frame(c(exp(fit$internal.summary.hyperpar$`0.5quant`),
                           exp(r_opt$par),
                           exp(tmb_opt$par)))

rownames(kable_data) <- c("$\\texttt{R-INLA}$", "R", "$\\texttt{TMB}$")
colnames(kable_data) <- c("Posterior mode")

@@

Compare the variances for $\log(\theta)$:

```{r}
sd_out$cov.fixed # TMB
fit$internal.summary.hyperpar$sd^2 # INLA
```

@@

### Is `h$fn` the same as in R?

@@

Could make a version of `h$fn` on the same scale as `post_marginal_theta` for a plot:
```{r results=FALSE}
tmb_post_marginal_theta <- function(theta) {
  de_nl(h$fn, x = log(theta))
} # (Have tested that this function does as it should)

```{r fig.cap = "At least this verifies that the optimisation is sensible."}
# Normalised dense_theta and tmb_dense_theta since they include different constants
ggplot(dense_theta) +
  geom_line(aes(x = input, y = norm_output), col = midblue) +
  geom_line(data = tmb_dense_theta, aes(x = input, y = norm_output), col = midpink) +
  theme(axis.text.y=element_blank()) +
  annotate("text", x = 0.1, y = 0.01, label = "TMB", col = midpink) +
  annotate("text", x = 0.045, y = 0.012, label = "R", col = midblue) +
  theme_minimal()
```

Does `TMB` do the inner optimisation differently?

@@

### Getting to the full conditional of $x$ in TMB

Probably going to need $Q^\star(\theta)$ ($n \times n$ precision matrix) and $\mu^\star(\theta)$ ($n \times 1$ mean vector).
For example, the most basic version of the full conditional would be the Gaussian approximation
\begin{equation}
\tilde p(x_i \, | \, \theta, y) = \mathcal{N}(x_i \, | \, \mu^\star_i(\theta), 1 / q^\star_i(\theta)).
\end{equation}

Precision matrix $Q^\star(\theta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\mu^\star(\theta)$, having $(i, j)$th entry
\begin{equation}
Q^\star(\theta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \theta, y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \mu^\star(\theta)}
\end{equation}

An obvious choice for finding the precision would be `h$he(tmb_opt$par)` but this gives the error "Hessian not yet implemented for models with random effects".
There may be another way to do it using `h$env$spHess()` (some kind of sparse Hessian) but if this is the case then it doesn't make sense to not have `h$he` just call the alternative way.
This matrix is of the class `dsCMatrix` which is for symmetric, sparse numeric matrices in the compressed column-oriented format.

`h$env$spHess` takes two arguments:

  1. `par` (default `h$env$par`) should be both the hyperparameters and random effects
  2. `random` (`TRUE` or `FALSE`)

and the code is given by:

```{r eval=FALSE}
function(par = obj$env$par, random = FALSE) {
  if (!random) {
    Hfull@x[] <- ev(par) # Whats ev?
    Hfull # H full <=> Full Hessian?
  }
  else if (skipFixedEffects) {
    .Call("setxslot", Hrandom, ev(par), PACKAGE = "TMB")
  }
  else {
    Hfull@x[] <- ev(par)
    Hfull[r, r]
  }
}
```

Set a range of test parameters `test_pars`:

```{r results=FALSE}
full_pars <- function(input) {
  setNames(input, c("x", "l_theta"))
}

par1 <- full_pars(c(0, 0))
par2 <- full_pars(c(0, -1))
par3 <- full_pars(c(-1, 0))

# Doing the inner loop manually
par4 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = 0), 0))
par5 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = -1), -1))

test_pars <- list(par1, par2, par3, par4, par5)
```

```{r}
lapply(test_pars, h$env$spHess)
```

`random = TRUE` seems behave in unpredictable ways e.g. always returning the value calculated last when `random = FALSE` (perhaps only when using `apply` functions and not in other situations, so results here omitted):

```{r results=FALSE}
lapply(test_pars, h$env$spHess, random = TRUE)
```

1) With `random = FALSE` the number changes as a function of the second argument, `l_theta` (the hyperparameters).
The Hessian is $2 \times 2$ with a dot ("structural zero") outside entry $(1, 1)$.
2) With `random = TRUE` the number doesn't change at all and it's now a $1 \times 1$ matrix.

Utilising the objective function $f$ which simply calculates the negative log posterior as a function of both random effects and hyperparameters:

```{r}
lapply(test_pars, f$he)
```

This produces the same answers as numerical differentiation of the same function (results omitted for space):

```{r results=FALSE}
lapply(test_pars, optimHess, fn = f$fn, gn = f$gn)
```

In this situation we know what the precision of $x$'s full conditional is (and its a function of $\theta$!):

```{r}
sapply(c(exp(0), exp(-1)), function(theta) inner_loop(theta)$tau_n)
```

Could look more at `sdreport` [here](https://github.com/kaskr/adcomp/blob/master/TMB/R/sdreport.R) see how these functions are used there (there are more comments on the Github page than when just looking via R).

```{r}
nl_full_conditional_x <- function(x, theta, log_input = FALSE) {

  if(log_input == TRUE) {
    theta <- exp(theta) # Convert from log(theta) to theta
  }

  par <- inner_loop(theta)

  return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE))
}
```

>>>>

Consider a three-stage model
\begin{alignat}{2}
  &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \, 0,  Q(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
\end{alignat}
with posterior
\begin{equation}
  p(x, \theta \, | \, y) \propto p(y, x, \theta) = p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta).
\end{equation}

@@

Approximate the posterior marginal $p(\theta \, | \, y) \propto L(\theta)$ using the Laplace approximation $L(\theta) \approx L^\star(\theta)$:
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta)}{\tilde p_G(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} \equiv L^\star(\theta)
\end{equation}

@@

<!-- *Note: Does `TMB` keep the $-\frac{n}{2} \log(2\pi)$ constant here $-\log L^\star_f(\theta) = -\frac{n}{2} \log(2\pi) + \frac{1}{2} \log \det (Q^\star(\theta)) + f(\mu(\theta), \theta)$?* -->

@@

We could also use an adaptive Gauss Hermite quadrature rule.
See the [Implementing Approximate Bayesian Inference using Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.

@@

This is the key innovation in the 2009 INLA paper.

@@

By construction, the distribution $p(x \, | \, \theta, y)$ is exact:

@@

Note that the code above in the section marked `(^)` is very similar to that in `blangiardo.cpp`.
Recall that the Laplace approximation $\tilde p(\theta \, | \, y)$, which in this instance coincides with the exact posterior, is given by
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(y, \mu(\theta), \theta)}{\det(Q(\theta))^{1/2}}.
\end{equation}
This form can be seen in `(^)`:

* $p(y, \mu(\theta), \theta)$ corresponds to the calls to `dgamma`, `dnorm` and sum of `dnorm` evaluated at $\mu(\theta)$ which here is simply `par$x_n` (for the Normal distribution the mean of the distribution is the same as the mode).
* The logarithm of $1 / \det(Q(\theta))^{1/2}$ when $\theta$ is a scalar is given by `-0.5 * log(par$tau_n)`.

@@

```{r}
trapezoid_rule <- function(x, spacing) {
  0.5 * spacing * (2 * sum(x) - x[1] - x[2])
}

eval_grid <- function(grid = NULL, spacing = NULL, uniform = FALSE, K = NULL, min = NULL, max = NULL, f) {
  if(uniform) {
    grid <- seq(min, max, length.out = K)
    spacing <- (max - min) / K
  }
  df <- data.frame(input = grid, output = sapply(grid, f))
  df <- mutate(df, norm_output = output / trapezoid_rule(output, spacing))
  return(df)
}

blangiardo_theta <- eval_grid(
  uniform = TRUE,
  K = 25, min = 0.001, max = 0.3,
  f = post_marginal_theta
)

dense_theta <- eval_grid(
  uniform = TRUE,
  K = 500, min = 0.001, max = 0.3,
  f = post_marginal_theta
)
```

@@

inla_theta <- eval_grid(exp(points), f = post_marginal_theta, spacing = delta_z)

@@

# Different normalising constant?
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y = element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  theme_minimal()
```

@@

```{r}
h <- MakeADFun(
  data = dat,
  parameters = param,
  random = c("x"),
  DLL = "blangiardo",
  hessian = TRUE,
  silent = TRUE
)
```

@@

```{r}
f <- MakeADFun(
  data = dat,
  parameters = param,
  DLL = "blangiardo",
  hessian = TRUE,
  silent = TRUE
)
```

@@

Comparison of the posterior mode and standard deviation of $\log(\theta)$:

```{r}
#' For some reason the mode in the INLA result below is NA, how to fix this?
#' According to Havard, this is a feature not a bug, and he hasn't found a good way to compute the mode yet:
#' https://groups.google.com/g/r-inla-discussion-group/c/mbSsNO_bydo

kable_data <- data.frame(
  "Mode" = c(NA, r_opt$par, tmb_opt$par),
  "Variance" = c(fit$internal.summary.hyperpar$sd^2, sd_out$cov.fixed, NA)
)

rownames(kable_data) <- c("$\\texttt{R-INLA}$", "R", "$\\texttt{TMB}$")

@@

@@

Could make a version of `h$fn` on the same scale as `post_marginal_theta` for a plot:

```{r results=FALSE}
tmb_post_marginal_theta <- function(theta) {
  de_nl(h$fn, x = log(theta))
} # (Have tested that this function does as it should)

```{r fig.cap = "At least this verifies that the optimisation is sensible."}
# Normalised dense_theta and tmb_dense_theta since they include different constants
ggplot(dense_theta) +
  geom_line(aes(x = input, y = norm_output), col = midblue) +
  geom_line(data = tmb_dense_theta, aes(x = input, y = norm_output), col = midpink) +
  theme(axis.text.y = element_blank()) +
  annotate("text", x = 0.1, y = 0.01, label = "TMB", col = midpink) +
  annotate("text", x = 0.045, y = 0.012, label = "R", col = midblue) +
  theme_minimal()
```

### Unit tests {.tabset .tabset-fade}

#### That `h$fn` the same as in R

@@

#### That the inner optimisation is the same

@@

###

<!-- ### Getting to the full conditional of $x$ in TMB -->

<!-- Probably going to need $Q^\star(\theta)$ ($n \times n$ precision matrix) and $\mu(\theta)$ ($n \times 1$ mean vector). -->
<!-- For example, the most basic version of the full conditional would be the Gaussian approximation -->
<!-- \begin{equation} -->
<!-- \tilde p(x_i \, | \, \theta, y) = \mathcal{N}(x_i \, | \, \mu_i(\theta), 1 / q^\star_i(\theta)). -->
<!-- \end{equation} -->

<!-- Precision matrix $Q^\star(\theta)$ is the Hessian of the negative logarithm of the posterior evaluated at $\mu(\theta)$, having $(i, j)$th entry -->
<!-- \begin{equation} -->
<!-- Q^\star(\theta)_{i, j} = - \frac{\partial^{2} \log p(x \, | \, \theta, y)}{\partial x_{i} \partial x_{j}} \Big\rvert_{x = \mu(\theta)} -->
<!-- \end{equation} -->

<!-- An obvious choice for finding the precision would be `h$he(tmb_opt$par)` but this gives the error "Hessian not yet implemented for models with random effects". -->
<!-- There may be another way to do it using `h$env$spHess()` (some kind of sparse Hessian) but if this is the case then it doesn't make sense to not have `h$he` just call the alternative way. -->
<!-- This matrix is of the class `dsCMatrix` which is for symmetric, sparse numeric matrices in the compressed column-oriented format. -->

<!-- `h$env$spHess` takes two arguments: -->

<!--   1. `par` (default `h$env$par`) should be both the hyperparameters and random effects -->
<!--   2. `random` (`TRUE` or `FALSE`) -->

<!-- and the code is given by: -->

<!-- ```{r eval=FALSE} -->
<!-- function(par = obj$env$par, random = FALSE) { -->
<!--   if (!random) { -->
<!--     Hfull@x[] <- ev(par) # Whats ev? -->
<!--     Hfull # H full <=> Full Hessian? -->
<!--   } -->
<!--   else if (skipFixedEffects) { -->
<!--     .Call("setxslot", Hrandom, ev(par), PACKAGE = "TMB") -->
<!--   } -->
<!--   else { -->
<!--     Hfull@x[] <- ev(par) -->
<!--     Hfull[r, r] -->
<!--   } -->
<!-- } -->
<!-- ``` -->

<!-- Set a range of test parameters `test_pars`: -->

<!-- ```{r results=FALSE} -->
<!-- full_pars <- function(input) { -->
<!--   setNames(input, c("x", "l_theta")) -->
<!-- } -->

<!-- par1 <- full_pars(c(0, 0)) -->
<!-- par2 <- full_pars(c(0, -1)) -->
<!-- par3 <- full_pars(c(-1, 0)) -->

<!-- # Doing the inner loop manually -->
<!-- par4 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = 0), 0)) -->
<!-- par5 <- full_pars(c(get_tmb_inner_optimised_value(l_theta = -1), -1)) -->

<!-- test_pars <- list(par1, par2, par3, par4, par5) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- lapply(test_pars, h$env$spHess) -->
<!-- ``` -->

<!-- `random = TRUE` seems behave in unpredictable ways e.g. always returning the value calculated last when `random = FALSE` (perhaps only when using `apply` functions and not in other situations, so results here omitted): -->

<!-- ```{r results=FALSE} -->
<!-- lapply(test_pars, h$env$spHess, random = TRUE) -->
<!-- ``` -->

<!-- 1) With `random = FALSE` the number changes as a function of the second argument, `l_theta` (the hyperparameters). -->
<!-- The Hessian is $2 \times 2$ with a dot ("structural zero") outside entry $(1, 1)$. -->
<!-- 2) With `random = TRUE` the number doesn't change at all and it's now a $1 \times 1$ matrix. -->

<!-- Utilising the objective function $f$ which simply calculates the negative log posterior as a function of both random effects and hyperparameters: -->

<!-- ```{r} -->
<!-- lapply(test_pars, f$he) -->
<!-- ``` -->

<!-- This produces the same answers as numerical differentiation of the same function (results omitted for space): -->

<!-- ```{r results=FALSE} -->
<!-- lapply(test_pars, optimHess, fn = f$fn, gn = f$gn) -->
<!-- ``` -->

<!-- In this situation we know what the precision of $x$'s full conditional is (and its a function of $\theta$!): -->

<!-- ```{r} -->
<!-- sapply(c(exp(0), exp(-1)), function(theta) inner_loop(theta)$tau_n) -->
<!-- ``` -->

<!-- Could look more at `sdreport` [here](https://github.com/kaskr/adcomp/blob/master/TMB/R/sdreport.R) see how these functions are used there (there are more comments on the Github page than when just looking via R). -->

<!-- ```{r} -->
<!-- nl_full_conditional_x <- function(x, theta, log_input = FALSE) { -->

<!--   if(log_input == TRUE) { -->
<!--     theta <- exp(theta) # Convert from log(theta) to theta -->
<!--   } -->

<!--   par <- inner_loop(theta) -->

<!--   return(-dnorm(x, par$x_n, sqrt(1 / par$tau_n), log = TRUE)) -->
<!-- } -->
<!-- ``` -->
